---
title: "Discrete Event Simulation"
author: "Nicholas Head"
date: "February 18, 2015"
layout: default
---

```{r echo=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.path='{{ site.url }}/images/discrete-event-simulation-')
```

```{r echo=FALSE}
cumNumberTimeProduct <- 0          # cumulative number-time product
queueSizeTs <- 0         # plot instantaneous q size
timeDeltaTs <- 0         # plot time delta
plotSamples <- 100

LogInstrumentation <- function (simTime, numberInSystem) {
  if (simTime < plotSamples) { 
      queueSizeTs <<- append(queueSizeTs,numberInSystem)
      timeDeltaTs <<- append(timeDeltaTs,simTime) 
  }
  return(simTime)
}

IncrementPopulation <- function (numberInSystem, simTime, lastEventTime, increment=1) {
  cumNumberTimeProduct <<- cumNumberTimeProduct + numberInSystem * (simTime - lastEventTime)  
  numberInSystem <- numberInSystem + increment
  lastEventTime <- LogInstrumentation(simTime, numberInSystem)
  return (c(numberInSystem, lastEventTime))
}
```

###M/M/1 Queue

A single-server queue with a simple birth-death process:

$$\begin{array}{l l}
    \text{transition} & \quad \text{rate} \\
    i \rightarrow i + 1 & \quad \lambda & \quad  (i \geq 0) \\
    i \rightarrow i - 1 & \quad \mu     & \quad  (i \geq 1) \\
  \end{array} $$

The simulation loop:
```{r}
RunSim.MM1 <- function (lambda, mu, simDuration) {
  simTime <- 0    # sim time
  timeForNextBirth <- 0         # time for next arrival
  timeForNextDeath <- simDuration     # time for next departure
  lastEventTime <- simTime   # tmp var for last event time
  lastBusyTimeStart <- 0         # tmp var for last busy-time start
  numberInSystem <- 0          # number in system
  totalBusyTime <- 0          # total busy time
  totalCompletions <- 0          # total completions
  
  set.seed(1)
  
  while (simTime < simDuration) {
      if (timeForNextBirth < timeForNextDeath) {      # birth
          simTime <- timeForNextBirth
          
          ret = IncrementPopulation(numberInSystem, simTime, lastEventTime) # instrumentation
          numberInSystem = ret[1]
          lastEventTime = ret[2]
          
          timeForNextBirth <- simTime + rexp(1, lambda)
          
          if(numberInSystem == 1) { 
              lastBusyTimeStart <- simTime
              timeForNextDeath <- simTime + rexp(1, mu)  # exponential  interarrival period
          }
      } else {            # death
          simTime <- timeForNextDeath
          
          ret = IncrementPopulation(numberInSystem, simTime, lastEventTime, -1) # instrumentation
          numberInSystem = ret[1]
          lastEventTime = ret[2]
          totalCompletions <- totalCompletions + 1
          
          if (numberInSystem > 0) { 
              timeForNextDeath <- simTime + rexp(1, mu)  # exponential  service period
          }
          else { 
              timeForNextDeath <- simDuration
              totalBusyTime <- totalBusyTime + simTime - lastBusyTimeStart
          }
      }   
  }
  return (c(simTime, totalBusyTime, totalCompletions))
}
```
Kick off the simulation with appropriate parameters:

```{r}
lambda <- 0.7500188    # birth rate
mu <- 1.0000    # death rate
simDuration   <- 10^5 # duration of sim

ret = RunSim.MM1(lambda, mu, simDuration)
simTime = ret[1]
totalBusyTime = ret[2]
totalCompletions = ret[3]
```

Average busy period: $\frac{1}{\mu - \lambda}$
```{r}
1/(mu - lambda)
```

Average waiting time: $\frac{1}{\mu - \lambda} - \frac{1}{\mu} = \frac{\rho}{\mu - \lambda}$
```{r}
1/(mu - lambda) - 1/mu
```

Theoretical traffic intensity: $\rho = \frac{\text{arrival rate}}{\text{service rate}} = \frac{\lambda}{\mu}$

```{r}
lambda/mu
```


```{r echo=FALSE}
u <- totalBusyTime/simTime       # traffic intensity
N <- cumNumberTimeProduct/simTime       # mean queue length (see the Load Average notes)
x <- totalCompletions/simTime       # mean throughput C/T
r <- N/x             # mean residence time (from Little's law: Q = XR)
q <- sum(queueSizeTs)/max(timeDeltaTs) # estimated queue length for plot

library(knitr)
m <- as.matrix(t(c(u, N, x, r, q)))
colnames(m) = c("traffic intensity", "mean queue length", "traffic intensity", "mean residence time", "estimated queue length ")
kable(m)
plot(timeDeltaTs, queueSizeTs, type="s", xlab="Time",ylab="Instantaneous queue length",main="M/M/1 Simulation")

```

We can double check our theoretical results using the 'queueing' package:
```{r eval=TRUE}
library(queueing)

## create input parameters
i_mm1 <- NewInput.MM1(lambda=lambda, mu=mu, n=3)
## Build the model
summary(QueueingModel(i_mm1))
```


```{r echo=FALSE, eval=TRUE}
library(testthat)

epsilon <- 1e-6
expect_equal(simTime, simDuration) 
expect_equal(totalBusyTime, 74774.78, tolerance=epsilon)
expect_equal(cumNumberTimeProduct, 302188.1, tolerance=epsilon)
expect_equal(u, 0.7477478, tolerance=epsilon)
expect_equal(N, 3.021881, tolerance=epsilon)
expect_equal(x, 0.74954, tolerance=epsilon)
expect_equal(r, 4.031647, tolerance=epsilon)
```


###M/M/C Queue

Now we examine a queueing model with support for $c$ servers:

$$\begin{array}{l l}
    \text{transition} & \quad \text{rate} \\
    i \rightarrow i + 1 & \quad \lambda & \quad  (i \geq 0) \\
    i \rightarrow i - 1 & \quad \mu \text{ min} (i, c)     & \quad  (i \geq 1) \\
  \end{array} $$

The simulation loop is identical to the version for the M/M/1 queue apart from the modification to the timeToNextDeath reflecting the new transition logic:

```{r eval=FALSE}
timeForNextDeath <- simTime + rexp(1, min(numberInSystem, c) / mu)
```

```{r echo=FALSE}
RunSim.MMC <- function (lambda, mu, c, simDuration) {
  simTime <- 0    # sim time
  timeForNextBirth <- 0         # time for next arrival
  timeForNextDeath <- simDuration     # time for next departure
  lastEventTime <- simTime   # tmp var for last event time
  lastBusyTimeStart <- 0         # tmp var for last busy-time start
  numberInSystem <- 0          # number in system
  totalBusyTime <- 0          # total busy time
  totalCompletions <- 0          # total completions
  
  set.seed(1)
  
  while (simTime < simDuration) {
      if (timeForNextBirth < timeForNextDeath) {      # birth
          simTime <- timeForNextBirth
          
          ret = IncrementPopulation(numberInSystem, simTime, lastEventTime) # instrumentation
          numberInSystem = ret[1]
          lastEventTime = ret[2]
          
          timeForNextBirth <- simTime + rexp(1, lambda)
          
          if(numberInSystem == 1) { 
              lastBusyTimeStart <- simTime
              timeForNextDeath <- simTime + rexp(1, mu)  # exponential  interarrival period
          }
      } else {            # death
          simTime <- timeForNextDeath
          
          ret = IncrementPopulation(numberInSystem, simTime, lastEventTime, -1) # instrumentation
          numberInSystem = ret[1]
          lastEventTime = ret[2]
          totalCompletions <- totalCompletions + 1
          
          if (numberInSystem > 0) { 
              timeForNextDeath <- simTime + rexp(1, min(numberInSystem, c) * mu)  # exponential  service period
          }
          else { 
              timeForNextDeath <- simDuration
              totalBusyTime <- totalBusyTime + simTime - lastBusyTimeStart
          }
      }   
  }
  return (c(simTime, totalBusyTime, totalCompletions))
}
```
Kick off the simulation with appropriate parameters:

```{r}
c = 3 # number of servers
lambda <- 0.7500188    # birth rate
mu <- 1.0000    # death rate
simDuration   <- 10^5 # duration of sim

ret = RunSim.MMC(lambda, mu, c, simDuration)
simTime = ret[1]
totalBusyTime = ret[2]
totalCompletions = ret[3]
```

NB: I do not think these empirical stats are working properly for the M/M/C model due to limitations in the simulation instrumentation

```{r echo=FALSE}
u <- totalBusyTime/simTime       # traffic intensity
N <- cumNumberTimeProduct/simTime       # mean queue length (see the Load Average notes)
x <- totalCompletions/simTime       # mean throughput C/T
r <- N/x             # mean residence time (from Little's law: Q = XR)
q <- sum(queueSizeTs)/max(timeDeltaTs) # estimated queue length for plot

library(knitr)
m <- as.matrix(t(c(u, N, x, r, q)))
colnames(m) = c("traffic intensity", "mean queue length", "mean throughput C/T", "mean residence time", "estimated queue length ")
kable(m)
plot(timeDeltaTs, queueSizeTs, type="s", xlab="Time",ylab="Instantaneous queue length",main="M/M/C Simulation")

```

We can double check our theoretical results using the 'queueing' package:
```{r eval=TRUE}
library(queueing)

## create input parameters
i_mmc <- NewInput.MMC(lambda=lambda, mu=mu, n=3, c = c)
## Build the model
summary(QueueingModel(i_mmc))
```


```{r echo=FALSE, eval=FALSE}
library(testthat)

epsilon <- 1e-6
expect_equal(simTime, simDuration) 
expect_equal(totalBusyTime, 74774.78, tolerance=epsilon)
expect_equal(cumNumberTimeProduct, 302188.1, tolerance=epsilon)
expect_equal(u, 0.7477478, tolerance=epsilon)
expect_equal(N, 3.021881, tolerance=epsilon)
expect_equal(x, 0.74954, tolerance=epsilon)
expect_equal(r, 4.031647, tolerance=epsilon)
```
